---
title: "CopernicusDownloadR_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CopernicusDownloadR_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Introduction 

The `CopernicusDownloadR` package provides a seamless interface between R and the Copernicus Marine Service, simplifying the process of subsetting and downloading marine datasets. This package is designed for researchers and practitioners working with oceanographic data, offering easy access to products:

[Global Ocean Physics Reanalysis](https://doi.org/10.48670/moi-00021)

[Global Ocean Biogeochemistry Hindcast](https://doi.org/10.48670/moi-00019)

Using `CopernicusDownloadR`, you can configure your environment, authenticate with the Copernicus Marine Service, and retrieve data based on custom parameters such as geographic location, depth, and temporal range.

### Installation
To install the development version of `CopernicusDownloadR`, ensure you have the `devtools` package installed. Then, run the following:

```{r install-copernicus, eval=FALSE}
# Install devtools if necessary
install.packages("devtools")

# Install CopernicusDownloadR
devtools::install_github("elifrag/CopernicusDownloadR")
```

Once installed, load the package:

```{r}
library(CopernicusDownloadR)
```

### Setting Up the Environment

The package requires Python to interface with the Copernicus Marine Service. Use the `setupCopernicusEnvironment` function to configure your Python environment and authenticate your Copernicus account.

To get started, register for a free account with the Copernicus Marine Service:

[Register Here](https://data.marine.copernicus.eu/register)


Here’s an example setup:

``` r
# Authenticate and configure Python environment
cm <- setupCopernicusEnvironment("your_username", "your_password")
```

The `cm` object returned will be used for subsequent data extraction functions.

### Listing Available Variables and Datasets

The `listVariables` function allows you to explore the variables and datasets available in the Copernicus Marine Service. You can request a complete list or just the names of the variables.

``` r
# List all variables with details
listVariables(show = "completeList")

# List only variable names
listVariables()
```

This function helps you identify which variables and datasets suit your research needs.

### Subsetting and Downloading Data

The `extractData` function allows you to download environmental data based on your specifications. You can define variables, geographic ranges (longitude, latitude, depth), time intervals, and frequency (daily or monthly). You can also define the path to the downloaded data and metadata information file. Here's how:

#### Example 1: Single Variable at a Point Location

``` r
# Download daily oxygen concentration for a specific location and time range
copernicusData <- extractData(
  cm = cm,
  variables = "oxygen",
  lon = -9.6467, lat = 36.7143,
  startDate = "2021-01-01", endDate = "2021-01-20",
  depth = c(10),
  frequency = "daily",
  data_output = "path_to_download_folder"
)
```

#### Example 2: Multiple Variables, Broad Geographic Region and Multiple Depths

``` r
# Download monthly data for ocean temperature and oxygen over a region and multiple depths
copernicusData <- extractData(
  cm = cm,
  variables = c("oceanTemperature", "oxygen"),
  lon = c(-30, 10), lat = c(30, 45),
  startDate = "2020-03-01", endDate = "2020-06-20",
  depth = c(1, 30),
  frequency = "monthly",
  data_output = "path_to_download_folder"
)
```

The extractData function automatically selects the appropriate dataset based on the requested variables and saves metadata information.


### Visualizing and Transforming Downloaded Data

Once you have downloaded data using `extractData`, you can visualize and manipulate it in R. Below are examples for importing, plotting, and saving NetCDF data, which may be single-band (e.g., one depth and time) or multi-band (e.g., multiple depths or times).

#### Step 1: Load and Inspect a NetCDF File

 Start by setting the path to your NetCDF file and inspecting its contents using the `ncdf4` package:

``` r
# Load the ncdf4 package
library(ncdf4)

# Set the path for the NetCDF file
# Replace "path_to_data_folder" with the file path to your downloaded dataset
ncfile <- "path_to_data_folder"

# Open the NetCDF file
nc <- nc_open(ncfile)

# Print information about the NetCDF file
print(nc)
```

This step allows you to examine the structure of the NetCDF file, including variables, dimensions (e.g., time, depth), and attributes.

#### Step 2: Importing and Visualizing Single-Band NetCDF Files

For single-band NetCDF files, you can load and visualize the data using the `terra` package:

``` r
# Load the terra package
library(terra)

# Import a single-band NetCDF file as a raster
single <- rast(ncfile)

# Print a summary of the raster
print(single)

# Plot the raster data
plot(single, main = "Single-Band NetCDF Raster")
```

#### Step 3: Importing and Visualizing Multi-Band NetCDF Files

For multi-band NetCDF files (e.g., multiple depths or times), you can also use the `rast` function:

``` r
# Import a multi-band NetCDF file
multi <- rast(ncfile)

# Print a summary of the multi-band raster
print(multi)

# Plot the first band (e.g., depth level or first time step)
plot(multi[[1]], main = "Multi-Band NetCDF Raster: Band 1")
```

#### Step 4: Selecting and Visualizing Specific Time Steps

To extract and visualize a specific time step from a multi-band NetCDF file:

``` r
# Select a specific time step (e.g., time step = 5)
time_step <- multi[[5]]

# Plot the selected time step
plot(time_step, main = "Time Step 5")

# Print a summary of the selected time step
print(time_step)
```

#### Step 5: Exporting NetCDF Data to raster

Save specific layers as GeoTIFF files:

``` r
# Save the selected raster layer as a GeoTIFF file
writeRaster(time_step, filename = "oxygen_concentration_time5.tif", overwrite = TRUE)
```

### Author

Eliza Fragkopoulou  
<https://github.com/elifrag>
### License

This package is licensed under the MIT License. See the
[LICENSE](LICENSE) file for more information.
